Spitzer, Gaia, and the Potential of the Milky Way

### 2013-09-27 text/paper1/ms.tex
remove sgr orbit from fig 3 and update caption

Sgr stream data has inspired a rich set of  models 
\citep[e.g.,][]{johnston99b, fellhauer06}.
Most recently, \citet[][hereafter LM10]{law10} have combined all
current data on the Sgr debris to constrain both a model of its evolution
and the potential in which it orbits.\footnote{However, new observational work by 
\cite{belokurov13} suggest that the trailing tail of Sgr debris does not
match the LM10 model.}
Figure \ref{fig:lm10} shows particle positions
from the final time-step of the LM10 N-body
simulation of dwarf satellite disruption along the expected Sgr orbit
in the best-fitting Milky Way halo model. The simulation was
run in a three-component potential, with a triaxial, logarithmic halo
model of the form
\begin{equation}
  \Phi_{halo} = v_{halo}^2 \ln(C_1 x^2 + C_2 y^2 + C_3 xy + (z/q_z)^2 + R_c^2)
\end{equation}
where $C_1$, $C_2$, and $C_3$ are combinations of the $x$ and $y$ axis
ratios ($q_1$, $q_2$) and orientation of the halo with respect to the
baryonic disk ($\phi$):
\begin{align}
  C_1 &= \frac{\cos^2\phi}{q_1^2} + \frac{\sin^2\phi}{q_2^2}\\
  C_2 &= \frac{\sin^2\phi}{q_1^2} + \frac{\cos^2\phi}{q_2^2}\\
  C_3 &= 2\sin\phi\cos\phi \left(q_1^{-2} - q_2^{-2}\right).
\end{align}

@@

LM10 demonstrate how a comparison of such simulations with the
observed debris enables a detailed reconstruction of Sgr's history:
its mass and stellar populations, orbit, rate of destruction, and even
original distribution in populations. 
This comparison enabled an assessment of the three-dimensional 
mass distribution of the Milky Way's dark matter
halo through the constraints on the potential parameters $v_{\rm halo}$, $q_1$, $q_z$, and $\phi$.

Combined {\it Spitzer} and {\it Gaia} measurements of distances and proper motions
of RR Lyrae in the Sgr debris will open up new avenues of
interpretation.

Figure~\ref{fig:gaia_errors} shows that a 2\% distance error is smaller than the 
distance range in the stream (top panel). Similarly, {\it Gaia} proper motion
error estimates correspond to tangential velocity errors less than the velocity
dispersion for much of the stream (bottom panel). The next section outlines a 
new method to take advantage of this information.

@@

\begin{figure}[h]
\begin{center}
%\includegraphics[width=0.47\textwidth,trim=5 0 0 0, clip]{/Users/adrian/projects/streams/plots/paper1/ps_distance.eps}
\includegraphics[width=0.47\textwidth,trim=5 0 0 0, clip]{/Users/adrian/projects/streams/plots/paper1/ps_distance.pdf}
\caption{Phase space distance ($D_{\rm ps}$) for 10 randomly selected stars integrated backwards in the correct potential (top) and a potential where $q_z$ is 25\% larger (bottom). The same 10 particles are used in both figures, so the initial conditions are identical. Red (thick, solid line) shows the galactocentric distance of the progenitor (in arbitrary, normalized units) over the integration. Horizontal (dashed) line shows $D_{\rm ps}=\sqrt{2}$.}\label{fig:ps_distance}
\end{center}
\end{figure}

@@

\begin{figure}[h]
\begin{center}
%\includegraphics[width=0.45\textwidth]{/Users/adrian/projects/streams/plots/paper1/objective_function.eps}
\includegraphics[width=0.45\textwidth]{/Users/adrian/projects/streams/plots/paper1/objective_function.pdf}
\caption{ 1D slices of the objective function (generalized variance) for each halo potential parameter. Curves are normalized to have the same value at the true parameter value (1.00), and show the effect of varying each parameter by $\pm10$\%. }\label{fig:objective}
\end{center}
\end{figure}

@@

It is also important to characterize the the limitations of this
method. Firstly, the measurement errors for RR Lyraes associated with
the very coldest streams \citep[e.g., the globular clusters Pal5 and
  GD1;][]{odenkirchen02,koposov10} will likely be too large to resolve
the minute differences in orbital properties between the debris and
satellite. Second, the current version of the algorithm
relies on accurate knowledge of the current position and velocity of
the parent satellite, which may not be available 
\citep[for example, the Orphan Stream;][]{belokurov07}. Lastly, the
present prescription neglects orbital evolution (e.g.,
dynamical friction) and N-body interactions which may lead to biases in
the estimated potential parameters.

>>>>

Sgr stream data has inspired a rich set of  models 
\citep[e.g.,][]{johnston99b, fellhauer06}.
Most recently, \citet[][hereafter LM10]{law10} combined all
the (then) current data on the Sgr debris to constrain both a model of its evolution
and the potential in which it orbits. (Note, new observational work by 
\cite{belokurov13} suggest that the trailing tail of Sgr debris does not
match the LM10 model.)
Figure \ref{fig:lm10} shows particle positions
from the final time-step of the LM10 N-body
simulation of dwarf satellite disruption along the expected Sgr orbit
in the best-fitting Milky Way halo model. The simulation was
run in a three-component potential, with a triaxial, logarithmic halo
model of the form
\begin{equation}
  \Phi_{halo} = v_{halo}^2 \ln(C_1 x^2 + C_2 y^2 + C_3 xy + (z/q_z)^2 + R_c^2)
\end{equation}
where $C_1$, $C_2$, and $C_3$ are combinations of the $x$ and $y$ axis
ratios ($q_1$, $q_2$) and orientation of the halo with respect to the
baryonic disk ($\phi$):
\begin{align}
  C_1 &= \frac{\cos^2\phi}{q_1^2} + \frac{\sin^2\phi}{q_2^2}\\
  C_2 &= \frac{\sin^2\phi}{q_1^2} + \frac{\cos^2\phi}{q_2^2}\\
  C_3 &= 2\sin\phi\cos\phi \left(q_1^{-2} - q_2^{-2}\right).
\end{align}

@@

%LM10 demonstrate how a comparison of such simulations with the observed debris enables a detailed reconstruction of Sgr's history: its mass and stellar populations, orbit, rate of destruction, and even original distribution in populations. 
A comparison of simulations and data enabled LM10 to make an assessment of the three-dimensional 
mass distribution of the Milky Way's dark matter
halo through the constraints on the potential parameters $v_{\rm halo}$, $q_1$, $q_z$, and $\phi$. Combined {\it Spitzer} and {\it Gaia} measurements of distances and proper motions
of RR Lyrae in the Sgr debris will open up new avenues for potential constraints. Figure~\ref{fig:gaia_errors} shows that a 2\% distance error is smaller than the 
distance range in the stream (top panel). Similarly, {\it Gaia} proper motion
error estimates correspond to tangential velocity errors less than the velocity
dispersion for much of the stream (bottom panel). The next section outlines a 
new method to take advantage of this information.

@@

\begin{figure}[h]
\begin{center}
%\includegraphics[width=0.47\textwidth,trim=5 0 0 0, clip]{/Users/adrian/projects/streams/plots/paper1/ps_distance.eps}
\includegraphics[width=0.47\textwidth,trim=5 0 0 0, clip]{/Users/adrian/projects/streams/plots/paper1/ps_distance.pdf}
\caption{Phase space distance ($D_{\rm ps}$) for 10 randomly selected stars integrated backwards in the correct potential (top) and a potential where $q_z$ is 25\% larger (bottom). The same 10 particles are used in both figures, so the initial conditions are identical. Horizontal (dashed) line shows $D_{\rm ps}=2$, for reference.}\label{fig:ps_distance}
\end{center}
\end{figure}

@@

\begin{figure}[h]
\begin{center}
%\includegraphics[width=0.45\textwidth]{/Users/adrian/projects/streams/plots/paper1/objective_function.eps}
\includegraphics[width=0.45\textwidth]{/Users/adrian/projects/streams/plots/paper1/objective_function.pdf}
\caption{ 1D slices of the objective function (generalized variance) for each halo potential parameter. The parameter values are normalized by the true values show the effect of varying each parameter by $\pm10$\%. The values of the objective function (vertical axis) are not interesting but note the minima around the truth (1.0).}\label{fig:objective}
\end{center}
\end{figure}

@@

It is also important to characterize the the limitations of this
method. Firstly, the measurement errors for RR Lyraes associated with
the very coldest streams \citep[e.g., the globular clusters Pal5 and
  GD1;][]{odenkirchen02,koposov10} will likely be too large to resolve
the minute differences in orbital properties between the debris and
satellite. Second, the present prescription neglects orbital evolution (e.g.,
dynamical friction) and scattering of stream stars due to the potential
of the satellite itself. Preliminary simulations (to be fully explored in 
forthcoming work) suggest that these first two points can be neglected 
for satellite masses between $\sim$$10^7$ and $\sim$$10^9~\mathrm{M}_{\odot}$. Lastly, the 
current version of the algorithm
relies on accurate knowledge of the current position and velocity of
the parent satellite, which may not be available 
\citep[for example, the Orphan Stream;][]{belokurov07}.
